```{r include = FALSE}
library(MASS)
library(corrplot)
library(tidyverse)
library(tidyr)
library(dplyr)
library(plyr)
library(GGally)
library(FactoMineR)
library(factoextra)
```

# Dimensionality reduction techniques

The theme of the fifth week was dimensionality reduction techniques.

##Data 

Some wrangling was done last week so I continued that one in separate file. 

```{r include = FALSE}
human_data <- read.csv("C:/Users/Tuukka/Desktop/IODS-project/data/humandata.csv")
rownames(human_data) <- human_data$X
human_data <- dplyr::select(human_data, -X)
```

Used dataset contains two different datasets that have been merged to one dataset _human\_data_. In the dataset is nine variables that contains information from Human Development Index and Gender Inequality Index. There is for example variables for ratios (female/male) about labour and education, life expectancy, maternal mortality ratio, adolescent birth rate and demale and share of parliamentary seats. Data is from year 2015. Dataset has 155 observations. 

```{r}
summary(human_data)
str(human_data)
ggpairs(human_data)
```
Ratios with edu and lab are numerical and their means are less than 1, but there is countries where ratios are more than 1 (portition of females are larger). GNIs are very different, but that is expected. Also maternal mortality ratios are very different between different countries. 

There is also some correlations. There are high correlations between for example maternal mortality ratio and life expectancy; life expectancy and expected years of education; adolescent birth rate and maternal mortality ratio.

It seems that most of countries are poor one (or there is a few very rich country). Values are distributed very differently between different variables. 

##Principal component analysis

```{r}
pca_human_nonSTD <- prcomp(human_data)
summary_pca_nonSTD <- summary(pca_human_nonSTD)
pca_pr_nonSTD <- round(1*summary_pca_nonSTD$importance[2, ], digits = 3) * 100
print(pca_pr_nonSTD)
pc_lab_nonSTD <- paste0(names(pca_pr_nonSTD), " (", pca_pr_nonSTD, "%)")
biplot(pca_human_nonSTD, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab_nonSTD[1], ylab = pc_lab_nonSTD[2])
```

```{r}
human_data_STD <- scale(human_data)
pca_human_STD <- prcomp(human_data_STD)
summary_pca_STD <- summary(pca_human_STD)
pca_pr_STD <- round(1*summary_pca_STD$importance[2, ], digits = 3) * 100
print(pca_pr_STD)
pc_lab_STD <- paste0(names(pca_pr_STD), " (", pca_pr_STD, "%)")
biplot(pca_human_STD, cex = c(0.8, 1), col = c("grey40", "red"), xlab = pc_lab_STD[1], ylab = pc_lab_STD[2])
```

Before standardization biplot is quite different than after standardization. It seems that GNI affects more than other variables so rich countries are on the right-hand side. Without standardization the first component explains 100 % of variance. After scailing the first component captures 54 % and the second one 16 % of the variace. The reason for the difference is that the PCA keeps all variables equal. That means that if value (mean) is larger than others, that variable affects more to the PCA. 

Bacause adding captions to the messy plot would make it even more messier, I will add some notations in the text. There is pointing two dimensions downwards. Those represent share of women in the economy and in the parliament. Bigger shares will affect to the PCA2 so that they make its value lower. To the right (higher values in PC1) is about education and life expectancy so more about the standards of living. Lower values in PC1 corresponds to mortality rate and birthrate.

PC2 is about the meaning and share of women in the economy and society. There is much differences between countries in PC2: it seems that european countries gets quite low values in PC2. Some African (southern part) countries are on the same level with western countries. Islamic countries have higher values and make cluster. PC1 is about the quality of life, education and birth rates. Western countries get better values because education levels and longer life expectancy. African countries are on the right-hand side because high birth rate and maternal mortality rate. 

##Multiple Correspondence Analysis
For the multiple correspondence analysis is needed a new dataset. I choose only some columns. The used dataset is a survey about tea. It has 300 observations in 36 variables. 

```{r include = FALSE}
data(tea)
keep_columns <- c("Tea", "How", "pub", "dinner", "resto", "price", "tearoom")
tea_data <- dplyr::select(tea, one_of(keep_columns))
str(tea_data)
summary(tea_data)
gather(tea_data) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

I kept seven variables: tea variety, is something added to tea, is tea drunk in the pub, is tea drunk with dinner, is tea drunk in the restaurant, price of tea and is tea drunk in the tearooms. With ggpairs can the dataset be observed. There are some differences in distributions. With one sentence the dataset can be described that responders are drinking their no-cheap Earl Grey tea without lemon or milk and not with dinner nor in the pub nor tearoom nor restaurant. 

```{r}
mca_tea <- MCA(tea_data, ncp = 5, graph = FALSE)
summary(mca_tea)
plot(mca_tea, invisible=c("ind"), habillage = "quali")
fviz_contrib(mca_tea, choice = "var", axes = 1)
fviz_contrib(mca_tea, choice = "var", axes = 2)
fviz_contrib(mca_tea, choice = "var", axes = 3)
```

The MDA is done 14 dimensions but after first three dimensions explain more variance than others. With three dimensions 31 % of variance is explained. In dimension 1 are variables _resto_, _pub_ and _tearoom_, so places where to drink. In dimension 2 are variables _Tea_ and _Price_, so variables about the tea. In dimensions 3 are variables _How_ and _price_, so other two variables. eta2 table tells correlations between dimensions and variables. With three bar charts tell how different variables contribute to the dimensions. They tell same story than the eta-tables.

Factor map (biplot) is drawn. It is quite hard to interpret. Firstly we can see the "lonely rider", _other_ that indicates that tea is drunk with something else than milk or lemon (that are nearly on origo). Black and green tea are drunk with dinner. Most of points are on the origo so their contributions to the dimensions 1 and 2 are quite low. 